---
title: "Lab 1"
output:
  html_document: default
  pdf_document: default
date: "2023-09-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Math for Data Science: Lab 1

## Warm Up

1.  R runs on a few different kinds of objects; let's get familiar with them.
    a.  **Vectors:** Create a vector containing only the number 1. Create an empty vector of length 5. Create a vector with the numbers 1, 2, 3, 6, 5.
    b.  **Data Frames:** Create an empty data frame with columns called x and y, both of length 5.

```{r}

vector_1 <- c(1)
  
empty_vector <- c(NA, NA, NA, NA, NA)
  
vector_2 <- c(1,2,3,6,5)

data_frame1 <- data.frame(x = 5, y = 5)

vector_1
empty_vector
vector_2
data_frame1
```

*2. Learn how to index and summarize your data.*

a\. Create a vector called nums containing the numbers 1, 6, 7, 8, 9.

b\. Grab the second element of this vector.

c\. Grab the second and third elements of this vector.

d\. Grab all elements of this vector except the first.

e\. Take the mean, minimum, maximum, and sum of this vector.

f\. Find the proportion of elements of this vector that are greater than *5*.

```{r}
nums <- c(1,6,7,8,9) #a. Create a vector called nums containing the numbers 1, 6, 7, 8, 9.

nums[2] #b. Grab the second element of this vector. 
nums[2:3]
nums[2:5] #d. Grab all elements of this vector except the first. 

mean(nums)
max(nums)
sum(nums)

mean(nums > 5) #f. Find the proportion of elements of this vector that are greater than 5. 
```

*3. Try out some basic probabilistic functions and operations.*

a\. Compute 5! using the factorial function.

```{r}
factorial(5)
```

b\. Compute (5 4) using the choose function.

```{r}
choose(5,4)
```

c\. Use the sample function to do some random sampling, with and without replacement. Try to

provide your own sampling probabilities.

```{r}
?sample

sample(c(1:6), 5, replace = TRUE)
sample(c(1:6), 5, replace = FALSE)

```

d\. Roll a die. Roll two fair dice. Roll a loaded die.

```{r}
roll_1 <- sample(1:6, 1, replace = TRUE)
roll_1

roll_2 <- sample(1:6, 2, replace = TRUE) #Replace because you want all the numbers to still be available
roll_2

loaded <- c(0.04, 0.2, 0.3, 0.4, 0.03, 0.03)
loaded_roll <- sample(1:6, 1, replace = TRUE, prob = loaded)
loaded_roll
```

e\. Learn about the set.seed command and try it out with your sampling. What does it do?

***Note: Whenever you run a sample, set a seed so that you can go back to the previous seed and get the exact sample back. "For probabilistic stuff, don't need for deterministic"***

```{r}
set.seed(123)
sample(1:6, 6, replace = TRUE)
sample(1:6, 6, replace = TRUE)

set.seed(123)
sample(1:6, 6, replace = TRUE) #the output of this should be the same as the output of the first sample.
```

------------------------------------------------------------------------

## Simulation

4\. We will now learn how to run a simulation. Recall the birthday problem from class. We will simulate

this to see if the mathematical result we showed in class holds up.

a\. Set a seed for reproducibility.

```{r}
set.seed(127)
```

b\. Start by creating one simulated class of 35 people. We will need a vector of randomly sampled

birthdays. Assume that all birthdays occur with equal probability and assume away leap years.

You can just label the days of the year from 1 to 365.

```{r}
## Sample from days of year (1:365), and select 35 students, WITH replacement because students can have the same birthday.
birth_days <- sample(1:365, 35, replace = TRUE)

birth_days
```

c\. Are there any repeat birthdays in your class? What are they?

```{r}
duplicates <- birth_days[duplicated(birth_days)] ## Return values where duplicated is TRUE
duplicates
```

d\. Now we will walk step by step through running a simulation. In this simulation, we will randomly

generate 100,000 fake classes and compute the average probability of having more than one of the

same birthday in a class (the frequentist interpretation of probability). First, generate an empty

vector called match of length 100,000 to hold our results.

```{r}
set.seed(100)

n <- 100000
matches <- rep(NA, n) # Empty list of size n (100,000)
```

e\. Write a loop that generates a fake class and saves the number of repeated birthdays in that class

into your empty vector.

```{r}
for (i in 1:n) {
  course <- sample(1:365, 35, replace = TRUE) # Sets 100,000 sets of 35 people classes
  matches[i] <- sum(duplicated(course))
}
```

f\. Summarize that output vector: what proportion of simulated classes have matching birthdays?

```{r}
prop.table(table(matches == 0))

##81 perent of the birthdays match
```

g\. Let's compare our simulated result to the mathematical result from class (see slide 30). Compute

the probability of matching birthdays using the formula we derived. (Note: the prod function

may be helpful here.)

```{r}
k <- 35
formula_result <- 1 - prod((365 - k + 1): 365)/365^k 

formula_result
0.81584/formula_result  # Formula result is only about 1 percent off the test result.
```

------------------------------------------------------------------------

## Plotting

*5. Now let's learn how to make plots. We'll recreate the plot from slide 31.*

a\. First, write a loop to compute the probability of a birthday match for k = 1 to 100 people in the

class. This time, let's save the output into a data frame with two variables, one for the x axis (k)

and another for the y axis (probability of birthday match).

```{r}
# create data frame to save output
to_plot <- data.frame(x = c(1:100), y = NA)

#loop to compute probability
for (k in c(1:100)) {
  to_plot$y[k]<- 1 - prod((365-k+1):365)/365^k
  }
```

b\. Use R's native plot function to make a quick plot.

```{r}
plot(to_plot$x, to_plot$y)
```

c\. This is ok, but we can do better. Install the ggplot package by typing install.packages("ggplot2")

into your console. Load this package with the library(ggplot2) command.

```{r}
install.packages("ggplot2")
library(ggplot2)
```

d\. Use ggplot to create a nicer plot, with axis labels, a title, and a red horizontal line at 0.5.

```{r}
ggplot(data = to_plot, aes(x = x, y = y)) +
geom_point() +
geom_hline(yintercept = 0.5, color = "red") +
xlab("k") +
ylab("probability of birthday match") +
ggtitle("The Birthday Problem") +
theme_minimal()
```

------------------------------------------------------------------------

## The Drunken Passenger Problem

*6. This is a common question in quantitative interviews. There are 100 passengers lined up to board an*

*airplane with 100 seats. Each seat has been assigned by the airline to one of the passengers. The first*

*passenger drunkenly sits down in a random seat. Each subsequent passenger takes their assigned seat*

*if it's available, and otherwise sits in a random available seat. What is the probability that the last*

*passenger in line gets to sit in their assigned seat?*

a\. We will run another simulation to answer this question. Begin by creating a data frame with three

columns: passenger which lists the passengers from 1 to 100; assigned which gives them their

assigned seats from the airline; and seat, representing the seats the passengers end up in, which

will begin by being empty and get filled over the simulation.

```{r}
#a. create data frame
df <- data.frame(passenger = c(1:100), assigned = sample(100, 100, replace = FALSE))

df$seat <- NA

```

b\. Let the first passenger pick a random seat. Fill in this value for seat in your data frame.

```{r}
# b. first passenger picks random seat
df$seat[1] <- sample(df$assigned, 1)

```

c\. Now fill in the second passenger's seat. Do this step by step. First create a vector of available

seats for the second passenger. Then give passenger 2 a seat using an ifelse statement.

```{r}
# c. second passenger's seat
free_seats <- df$assigned[!df$assigned %in% df$seat]
df$seat[2] <- ifelse(df$assigned[2] %in% df$seat, sample(free_seats, 1), df$assigned[2])

```

d\. With the same logic/syntax, write a loop assigning seats for passengers 3 through 100.

```{r}
# d. loop for all subsequent passengers
for (i in c(3:100)) {
df$seat[i] <- ifelse(df$assigned[i] %in% df$seat, sample(free_seats, 1), df$assigned[i])
}
```

2

e\. Now, wrap parts a-d up into a function that returns TRUE if the last passenger gets their assigned

seat and FALSE otherwise. Run this function 10,000 times using the replicate function. What

seems to be the probability of the last passenger getting their assigned seat?

```{r}
# e. now put this into a function
sim_plane <- function() {
df <- data.frame(passenger = c(1:100), assigned = sample(100, 100, replace = FALSE))
df$seat <- NA
df$seat[1] <- sample(df$assigned, 1)
for (i in c(2:100)) {
free_seats <- df$assigned[!df$assigned %in% df$seat]
df$seat[i] <- ifelse(df$assigned[i] %in% df$seat, sample(free_seats, 1), df$assigned[i])
}
success <- df$assigned[100]==df$seat[100]
return(success)
}
out <- replicate(10000, sim_plane())

```

f\. Adapt your function so that it accommodates an arbitrary number of passengers. Run it for a few

different values for the number of passengers. Does the number of passengers matter?

```{r}
# f. adapt for arbitrary number of passengers
sim_plane <- function(n) {
df <- data.frame(passenger = c(1:n), assigned = sample(n, n, replace = FALSE))
df$seat <- NA
df$seat[1] <- sample(df$assigned, 1)
for (i in c(2:n)) {
free_seats <- df$assigned[!df$assigned %in% df$seat]
df$seat[i] <- ifelse(df$assigned[i] %in% df$seat, sample(free_seats, 1), df$assigned[i])
}
success <- df$assigned[n]==df$seat[n]
return(success)
}
out <- replicate(10000, sim_plane(n = 50))
prop.table(table(out))

out <- replicate(10000, sim_plane(n = 100))
prop.table(table(out))

out <- replicate(10000, sim_plane(n = 500))
prop.table(table(out))
```

g\. Is there a mathematical/logical way to solve this problem? What is the intuition? (See solution to

Exercise 59 in Blitzstein and Hwang.)
